---
title: "730 Group Project"
author: "Rebekah Kristal with collaborators Amani Chehimi & Shane Fitzgerald"
date: "2024-11-30"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggplot2)
library(brms)
library(posterior)
library(tidybayes)
library(rstan)
library(bayesplot)
library(rstanarm)
library(loo)
```
# Data reformatting
```{r}
newdata <- read_csv("FreqCategories.csv")
newdata<-mutate(newdata, REGION=as.factor(REGION)) %>% mutate(AgeCat=as.factor(AgeCat)) %>% mutate(SEX=as.factor(SEX)) %>% mutate(RACENEW=as.factor(RACENEW))%>%mutate(EduCat=as.factor(EduCat)) %>% mutate(POORYN=as.factor(POORYN))

#converting y's into factor variable, changing range from 0-8 to 1-9 to match with model output
newdata1<-mutate(newdata, y=y+1) %>% mutate(y, factor(y, ordered=TRUE)) 
```
# Model fitting

## Amani's model: weighted linear regression with hierarchical variable
```{r}
modALoo <- brm(
  y | weights(Freq) ~ (1 | REGION + AgeCat + SEX + RACENEW + EduCat + POORYN), 
  data = newdata1,
  family = gaussian(), 
  iter = 1000, 
  chains = 4, 
  cores = getOption("mc.cores", 4), 
  seed = 12345
) 
```

## Shane's model: weighted ordinal regression
```{r}
modSLoo <- brm(
  y|weights(Freq)~REGION + AgeCat + SEX + RACENEW + EduCat + POORYN,
  data = newdata1, 
  family=cumulative(link="logit"),  
  iter = 1000, 
  chains = 4, 
  cores = getOption("mc.cores", 4), 
  seed = 12345
)
```

## Rebekah's model 1: weighted ordinal regression with hierarchical variable
```{r}
modRLoo1 <- brm(
  y | weights(Freq) ~ (1 | REGION + AgeCat + SEX + RACENEW + EduCat + POORYN),
  data = newdata1,
  family=cumulative(link="logit"),  
  iter = 1000, 
  chains = 4, 
  cores = getOption("mc.cores", 4), 
  seed = 12345
)
```

## Rebekah's model 2: weighted ordinal regression with interactions
```{r}
modRLoo2 <- brm(
  y|weights(Freq)~REGION + AgeCat + SEX + RACENEW + EduCat + POORYN + REGION*POORYN + REGION*RACENEW + RACENEW*POORYN, 
  data = newdata1,
  family=cumulative(link="logit"),  
  iter = 1000, 
  chains = 4, 
  cores = getOption("mc.cores", 4), 
  seed = 12345
)
```

# Our own PPC plots of expected vs observed counts

## get observed counts from data
```{r}
observed_counts <- dplyr::select(newdata1, c(y, Freq)) 
total_freq<-group_by(observed_counts, y) %>% summarise(total=sum(Freq)) 
observed_props<-mutate(total_freq, observed=total/sum(total)) %>% mutate(y=as.factor(y))


get_sum_stat<-function(y, row){(sum(y==5))/nrow(row)}

tobs<-observed_props[5,3]
```

## function to make ppc plot for each model of proportion of people in each response category
```{r}
make_ppc_plot <- function(model_name){
    predicted_catsR<-as.data.frame(posterior_predict(model_name))
    ynew_siR<-apply(predicted_catsR, 1, get_sum_stat, newdata)
    #ppc for proportion of observations in category 5
    hist(ynew_siR)
    abline(v = tobs)
    
    #ppc for all categories
    #formatting for ggplot
    posterior_preds_longR <- predicted_catsR %>%
      pivot_longer(cols = everything(), names_to = "chain", values_to = "predicted_category")
    
    posterior_preds_longR$predicted_category <- as.factor(posterior_preds_longR$predicted_category)
    
    category_countsR <- table(posterior_preds_longR$predicted_category)
    category_counts_dfR <- as.data.frame(category_countsR)
    colnames(category_counts_dfR) <- c("y", "Count")
    category_counts_propR<-mutate(category_counts_dfR, predicted=Count/(sum(Count)))
    
    combinedR<-left_join(observed_props, category_counts_propR, by="y")
    combined1R<-pivot_longer(combinedR, c(3,5), names_to = "Freq")
    
    #plot of proportion of each category for observed and predicted data
    ggplot(combined1R, mapping=aes(x=y, y=value, fill=Freq))+
      geom_bar(stat="identity", position="dodge")+
      labs(title = "Mental Health Category Proportions for Observed and Predicted Data",
           x = "Category",
           y = "Proportion") +
      theme_minimal() +
      theme(axis.text.x = element_text(angle = 45, hjust = 1))
}
```

## Make proportion PPC plots
```{r}
# Amani's model
make_ppc_plot(modALoo)
# Shane's model
make_ppc_plot(modSLoo)
# Rebekah's model 1
make_ppc_plot(modRLoo1)
# Rebekah's model 2
make_ppc_plot(modRLoo2)
```

# Model Comparisons using loo package
From module 11.2 code
*Error: Moment matching failed. Perhaps you did not set 'save_pars = save_pars(all = TRUE)' when fitting your model?*

```{r}
#Amani's model
looHier <- loo(modALoo, save_psis = TRUE) #, moment_match = TRUE)

#Shane's model
looOrd <- loo(modSLoo, save_psis = TRUE) #, moment_match = TRUE)

#Rebekah's model 1
looOrdHier <- loo(modRLoo1, save_psis = TRUE) #, moment_match = TRUE)

#Rebekah's model 2
looOrdInter <- loo(modRLoo2, save_psis = TRUE) #, moment_match = TRUE)
```
70/122/131/154 observations with a pareto_k > 0.7 in model. We recommend to set 'moment_match = TRUE' in order to perform moment matching for problematic observations. We recommend to run more iterations to get at least about 2200 posterior draws to improve LOO-CV approximation accuracy.

## PSIS Diagnostic Plots
```{r}
plot(looHier, diagnostic = c("k"), label_points = TRUE, main = "PSIS diagnostic plot")
plot(looOrd, diagnostic = c("k"), label_points = TRUE, main = "PSIS diagnostic plot")
plot(looOrdHier, diagnostic = c("k"), label_points = TRUE, main = "PSIS diagnostic plot")
plot(looOrdInter, diagnostic = c("k"), label_points = TRUE, main = "PSIS diagnostic plot")
```

## Marginal PPCs
```{r}
#Amani's model
yrep <- posterior_predict(modALoo)
ppc_loo_pit_qq(
  y = newdata1$y,
  yrep = yrep,
  lw = weights(looHier$psis_object)
)
#Shane's model
yrep <- posterior_predict(modSLoo)
ppc_loo_pit_qq(
  y = newdata1$y,
  yrep = yrep,
  lw = weights(looOrd$psis_object)
)
#Rebekah's model 1
yrep <- posterior_predict(modRLoo1)
ppc_loo_pit_qq(
  y = newdata1$y,
  yrep = yrep,
  lw = weights(looOrdHier$psis_object)
)
#Rebekah's model 2
yrep <- posterior_predict(modRLoo2)
ppc_loo_pit_qq(
  y = newdata1$y,
  yrep = yrep,
  lw = weights(looOrdInter$psis_object)
)
```

```{r}
yrep <- posterior_predict(modSLoo)
psis_all <- looOrd$psis_object
lw_all <- weights(psis_all) 
ppc_loo_pit_overlay(y = newdata1$y, yrep, lw = lw_all)

```

## Comparing Models on Expected Log Predictive Density
```{r}
#linear regression w region hierarchical variable vs ordinal regression w/ no hierarchy
loo_compare(looHier, looOrd)
#linear vs ordinal, both w region hierarchical
loo_compare(looHier, looOrdHier)
#ordinal without vs with interactions
loo_compare(looOrd, looOrdInter)
#ordinal without vs with hierarchical
loo_compare(looOrd, looOrdHier)
#ordinal with hierarchical no interactions vs interactions no hierarchical
loo_compare(looOrdHier, looOrdInter)
```

```{r}
loo_compare(looHier, looOrd, looOrdHier, looOrdInter)
```


```{r}
summary(modSLoo)
summary(modRLoo1)
```

```{r}
pp_check(modALoo)
```

```{r}
#reference category: mental health score = 0, region = 1 (Northeast), age = adult (18-34), sex = 1 (Male), race = 100 (white), Education = Grade 12 or less, no diploma, Poverty = 1 (At or above poverty threshold)
#           
ordModelCoeffs <- tribble(
            ~Coefficient, ~Estimate, ~Lower, ~Upper,
            "Intercept[1]", -0.35, -0.39, -0.31, 
            "Intercept[2]", 0.43, 0.39, 0.47,
            "Intercept[3]", 1.15, 1.11, 1.19, 
            "Intercept[4]", 1.56, 1.52, 1.60,
            "Intercept[5]", 2.10, 2.06, 2.14,
            "Intercept[6]", 2.54, 2.50, 2.59,
            "Intercept[7]", 3.07, 3.03, 3.12,
            "Intercept[8]", 3.68, 3.64, 3.74,
            "North Central/midwest", 0.04, 0.00, 0.07,
            "South", -0.07, -0.10, -0.04,
            "West", 0.09, 0.06, 0.12,
            "YoungAdult", -0.03, -0.06, -0.00,
            "Elderly", -0.38, -0.41, -0.36,
            "Female", 0.45, 0.43, 0.47,
            "Race Not Releasable", -0.53, -0.74, -0.32,
            "Black/African American", -0.41, -0.44, -0.38,
            "American Indian/Alaska Native", -0.10, -0.20, -0.00,
            "Asian", -0.54, -0.59, -0.50,
            "Multiple Race", 0.26, 0.20, 0.33,
            "High School Diploma or GED Graduate", 0.02, -0.01, 0.06,
            "Some College, AA, or Bachelor's Degree", 0.12, 0.09, 0.16,
            "Graduate, Professional, or other Degree", 0.13, 0.09, 0.17,
            "Below poverty threshold", 0.51, 0.48, 0.54)

# take eta = -beta
# ref ref ref = 1.56
# poverty ref ref = 1.56 - 0.51 = 1.05
# ref multiple ref = 1.56 - 0.26 = 1.30
# ref ref south = 1.56 - (-0.07) = 1.63
ordModelCoeffs

#reorder y axis
ordModelCoeffs$Coefficient <- factor(ordModelCoeffs$Coefficient, levels=unique(ordModelCoeffs$Coefficient))


ordModelCoeffs %>% ggplot(aes(y = Coefficient, x = Estimate, xmin = Lower, xmax = Upper)) + geom_pointinterval()
```


Rhat ~1, Bulk & Tail ESS all >4*100

# Equation for ordinal interactions
$ log\left(\frac{P(y_i \leq k)}{P(y_i > k)} \right) = \zeta_k - \eta_1x_{i1}-...-\eta_jx_{ij} - H_{I1}x_{i1}*x_{i2} - H_{I2}x_{i1}*x_{i3} - H_{I3}x_{i2}*x_{i3} + \varepsilon_i$

$ log\left(\frac{P(y_i \leq k)}{P(y_i > k)} \right) = above - H_{I1}Poverty*Region - H_{I2}Poverty*Race - H_{I3}Region*Race + \varepsilon_i$

# equation for ordinal hierarchical
$ log\left(\frac{P(y_i \leq k)}{P(y_i > k)} \right) = \zeta_k + \xi_{[i]k}- \eta_1x_{i1}-...-\eta_jx_{ij} + \varepsilon_i$


# amani's eq
$y_i = \alpha_{j[i]} + \beta^T*x_i + \varepsilon_i = M_i + \varepsilon_i$
$y_i = \mu_\alpha + \delta_{j[i]} + \beta^T*x_i + \varepsilon_i = \alpha_{j[i]} + \beta^T*x_i + \varepsilon_i = M_i + \varepsilon_i$
$y_i|(\alpha_{j[i]}, \beta, \sigma_y) \sim N(\alpha_{j[i]} + \beta^T*x_i, \sigma_y^2)$
$\alpha_{j[i]}|(\mu_\alpha, \sigma_\alpha) \sim N(\mu_\alpha, \sigma_\alpha^2),\ with\ priors\ on\ \beta, \mu_\alpha, \sigma_\alpha, \sigma_y$


# shane's eq
$ log\left(\frac{P(y_i \leq k)}{P(y_i > k)} \right) = \zeta_k - \eta_1x_{i1}-...-\eta_jx_{ij} + \varepsilon_i$

# estimates
$ log\left(\frac{P(y_i \leq 4)}{P(y_i > 4)} \right)$
$ P(y_i \leq 4)$

```{r}
newdata1 %>% group_by(POORYN, y) %>% summarise(count = sum(Freq)) %>% ggplot(aes(x = POORYN, y = y, size = count)) + geom_point()

```
# coeff plot using bayesplot (??)
```{r}
posterior <- as.array(modSLoo)
dim(posterior)
Coeffs <- dimnames(posterior)$variable[1:23]
mcmc_intervals(posterior, Coeffs)
mcmc_areas(posterior, Coeffs)
```

# get probas instead of cumulative log odds 
```{r}
invlogit <- function(x) 1/(1+exp(-x))
invlogit(1.56)
invlogit(1.05)
invlogit(1.30)
invlogit(1.63)
# take eta = -beta
# ref ref ref = 1.56
# poverty ref ref = 1.56 - 0.51 = 1.05
# ref multiple ref = 1.56 - 0.26 = 1.30
# ref ref south = 1.56 - (-0.07) = 1.63
```








