---
title: "730 Group Project"
author: "Rebekah Kristal with collaborators Amani Chehimi & Shane Fitzgerald"
date: "2024-11-30"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggplot2)
library(brms)
library(posterior)
library(tidybayes)
library(rstan)
library(bayesplot)
library(rstanarm)
library(loo)
```
# Data reformatting
```{r}
newdata <- read_csv("FreqCategories.csv") %>%  mutate(Weight = Freq / sum(Freq))
newdata<-mutate(newdata, weight.var=1/Freq) %>% mutate(REGION=as.factor(REGION)) %>% mutate(AgeCat=as.factor(AgeCat)) %>% mutate(SEX=as.factor(SEX)) %>% mutate(RACENEW=as.factor(RACENEW))%>%mutate(EduCat=as.factor(EduCat)) %>% mutate(POORYN=as.factor(POORYN))

#converting y's into factor variable, changing range from 0-8 to 1-9 to match with model output
newdata1<-mutate(newdata, y=y+1) %>% mutate(y, factor(y, ordered=TRUE)) 
```
# Model fitting

## Amani's model: weighted linear regression with hierarchical variable
```{r}
modALoo <- brm(
  y | weights(Weight) ~ (1 | REGION + AgeCat + SEX + RACENEW + EduCat + POORYN), 
  data = newdata1,
  family = gaussian(), 
  iter = 1000, 
  chains = 4, 
  cores = getOption("mc.cores", 4), 
  seed = 12345
) 
```

## Shane's model: weighted ordinal regression
```{r}
modSLoo <- brm(
  y|weights(Weight)~REGION + AgeCat + SEX + RACENEW + EduCat + POORYN,
  data = newdata1, 
  family=cumulative(link="logit"),  
  iter = 1000, 
  chains = 4, 
  cores = getOption("mc.cores", 4), 
  seed = 12345
)
```

## Rebekah's model 1: weighted ordinal regression with hierarchical variable
```{r}
modRLoo1 <- brm(
  y | weights(Weight) ~ (1 | REGION + AgeCat + SEX + RACENEW + EduCat + POORYN),
  data = newdata1,
  family=cumulative(link="logit"),  
  iter = 1000, 
  chains = 4, 
  cores = getOption("mc.cores", 4), 
  seed = 12345
)
```

## Rebekah's model 2: weighted ordinal regression with interactions
```{r}
modRLoo2 <- brm(
  y|weights(Weight)~REGION + AgeCat + SEX + RACENEW + EduCat + POORYN + REGION*POORYN + REGION*RACENEW + RACENEW*POORYN, 
  data = newdata1,
  family=cumulative(link="logit"),  
  iter = 1000, 
  chains = 4, 
  cores = getOption("mc.cores", 4), 
  seed = 12345
)
```

# Our own PPC plots of expected vs observed counts

## get observed counts from data
```{r}
observed_counts <- select(newdata1, c(y, Freq)) 
total_freq<-group_by(observed_counts, y) %>% summarise(total=sum(Freq)) 
observed_props<-mutate(total_freq, observed=total/sum(total)) %>% mutate(y=as.factor(y))


get_sum_stat<-function(y, row){(sum(y==5))/nrow(row)}

tobs<-observed_props[5,3]
```

## function to make ppc plot for each model of proportion of people in each response category
```{r}
make_ppc_plot <- function(model_name){
    predicted_catsR<-as.data.frame(posterior_predict(model_name))
    ynew_siR<-apply(predicted_catsR, 1, get_sum_stat, newdata)
    #ppc for proportion of observations in category 5
    hist(ynew_siR)
    abline(v = tobs)
    
    #ppc for all categories
    #formatting for ggplot
    posterior_preds_longR <- predicted_catsR %>%
      pivot_longer(cols = everything(), names_to = "chain", values_to = "predicted_category")
    
    posterior_preds_longR$predicted_category <- as.factor(posterior_preds_longR$predicted_category)
    
    category_countsR <- table(posterior_preds_longR$predicted_category)
    category_counts_dfR <- as.data.frame(category_countsR)
    colnames(category_counts_dfR) <- c("y", "Count")
    category_counts_propR<-mutate(category_counts_dfR, predicted=Count/(4000*5462))
    
    combinedR<-left_join(observed_props, category_counts_propR, by="y")
    combined1R<-pivot_longer(combinedR, c(3,5), names_to = "Freq")
    
    #plot of proportion of each category for observed and predicted data
    ggplot(combined1R, mapping=aes(x=y, y=value, fill=Freq))+
      geom_bar(stat="identity", position="dodge")+
      labs(title = "Mental Health Category Proportions for Observed and Predicted Data",
           x = "Category",
           y = "Proportion") +
      theme_minimal() +
      theme(axis.text.x = element_text(angle = 45, hjust = 1))
}
```

## Make proportion PPC plots
```{r}
# Amani's model
make_ppc_plot(modALoo)
# Shane's model
make_ppc_plot(modSLoo)
# Rebekah's model 1
make_ppc_plot(modRLoo1)
# Rebekah's model 2
make_ppc_plot(modRLoo2)
```

# Model Comparisons using loo package
Note: using brm to fit all models, then will "pass an array or matrix of log-likelihood values to the loo function," since stan_glm doesn't like the "|" hierarchical specification or the ordinal regression ("'family' must be one of binomial, gaussian, Gamma, inverse.gaussian, poisson, neg_binomial_2, mgcv::betar")

## get array/atrix of log-likelihood values to pass to the loo function

```{r}
#Amani's model
modALooAdj <- 
rel_n_eff <- relative_eff(exp(modALooAdj))
looHier <- loo(modALooAdj, r_eff = rel_n_eff, cores = 2, save_psis = TRUE)

#Shane's model
modSLooAdj <- 
rel_n_eff <- relative_eff(exp(modSLooAdj))
looOrd <- loo(modSLooAdj, r_eff = rel_n_eff, cores = 2, save_psis = TRUE)

#Rebekah's model 1
modRLoo1Adj <- 
rel_n_eff <- relative_eff(exp(modRLoo1Adj))
looOrdHier <- loo(modRLoo1Adj, r_eff = rel_n_eff, cores = 2, save_psis = TRUE)

#Rebekah's model 2
modRLoo2Adj <- 
rel_n_eff <- relative_eff(exp(modRLoo2Adj))
looOrdInter <- loo(modRLoo2Adj, r_eff = rel_n_eff, cores = 2, save_psis = TRUE)
```

## PSIS Diagnostic Plots
```{r}
plot(looHier)
plot(looOrd)
plot(looOrdHier)
plot(looOrdInter)
```

## Marginal PPCs
```{r}
#Amani's model
yrep <- posterior_predict(modALoo)
ppc_loo_pit_qq(
  y = newdata1$y,
  yrep = yrep,
  lw = weights(looHier$psis_object)
)
#Shane's model
yrep <- posterior_predict(modSLoo)
ppc_loo_pit_qq(
  y = newdata1$y,
  yrep = yrep,
  lw = weights(looOrd$psis_object)
)
#Rebekah's model 1
yrep <- posterior_predict(modRLoo1)
ppc_loo_pit_qq(
  y = newdata1$y,
  yrep = yrep,
  lw = weights(looOrdHier$psis_object)
)
#Rebekah's model 2
yrep <- posterior_predict(modRLoo2)
ppc_loo_pit_qq(
  y = newdata1$y,
  yrep = yrep,
  lw = weights(looOrdInter$psis_object)
)
```


## Comparing Models on Expected Log Predictive Density
```{r}
#linear regression w region hierarchical variable vs ordinal regression w/ no hierarchy
loo_compare(looHier, looOrd)
#linear vs ordinal, both w region hierarchical
loo_compare(looHier, looOrdHier)
#ordinal without vs with interactions
loo_compare(looOrd, looOrdInter)
#ordinal without vs with hierarchical
loo_compare(looOrd, looOrdHier)
#ordinal with hierarchical no interactions vs interactions no hierarchical
loo_compare(looOrdHier, looOrdInter)
```